hbm_paper_results

### 2024-01-02 paper_results.Rmd

# Setup 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(data.table, ggplot2, tidyverse, ggseg, ggsegExtra, paletteer, ggpubr, gridExtra, lemon, parallel, rstatix, lme4, interactions, ggrepel)
#Sys.setenv("MC_CORES"=10L)
options("mc.cores")

@@

```{r}
data_path <- "./derivatives"

@@

pheno_list.z <- paste0(pheno_list, ".z")
#also load raw data
ukb_cn_df <- fread("/Users/megardn/Desktop/BGD_Repos/combat_biovar/data/ukb_to_model/ukb_CN_data_simsites.csv")
```

@@

Within each combat config, test if there are significant differences in M and F subject's mean abs. centile error. Can test mean (not abs) centile errors, Z errors, etc, using `to_test` arg. FDR corrects for number of

@@

  #apply fdr correction
  n_comp=nrow(df)
  df.sex.t <- df.sex.t %>%
    dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = (n_comp*comp_multiplier)),
           sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                               p.val_fdr >= 0.05 ~ FALSE))
  return(df.sex.t)
  #detach()
}

@@

  #conduct t test
  df.lm <- df %>%
    group_by(dataset) %>%
    #returns values for last (highest order) predictor
    summarise(p.value = tail(tidy(lm(as.formula(f)))$p.value, n=1),
              estimate = tail(tidy(lm(as.formula(f)))$estimate, n=1),
              t.stat = tail(tidy(lm(as.formula(f)))$statistic[2], n=1),
              std.error = tail(tidy(lm(as.formula(f)))$std.error[2]), n=1) %>%
    ungroup()
  n_comp=nrow(df)

@@

##Load Data

pulled centile/z error calcs from cubic using `scp -r 'gardnerm@cubic-login.uphs.upenn.edu:/cbica/home/gardnerm/combat-biovar/ukb_basic/centile_csvs/*_diffs.csv' .`

pulled predicted centiles from cubic using `scp -r 'gardnerm@cubic-login.uphs.upenn.edu:/cbica/home/gardnerm/combat-biovar/ukb_basic/centile_csvs/*_predictions.csv' .`

@@

```{r}
combined_preds %>% 
  filter_at(vars(pheno_list), any_vars(. %in% c(1))) %>%
  dplyr::select(participant) %>%
  unique()
  dplyr::select(File_ID, participant, sex, dataset, pheno_list) %>%
  pivot_longer(cols=pheno_list, names_to="pheno", values_to="centile")
skimr::skim(qc_cent_df)

qc_cent_df %>%
  dplyr::filter(centile == 0)

qc_cent1_df <- qc_cent_df %>%
  dplyr::filter(centile == 1)

qc_cent1_df %>%
  ggplot() +
  geom_bar(aes(dataset, fill=participant)) +
  facet_wrap(~pheno)

ppt.list <- qc_cent1_df %>%
  dplyr::select(participant) %>%
  unique() %>%
  as.list() %>%
  unname() %>%
  unlist()

qc_z_df <- combined_preds %>% 
  filter_at(vars(pheno_list.z), any_vars(. %in% c(1, 0, Inf, -Inf, NA)))

unique(qc_z_df$participant)

qc_cent_df %>%
  dplyr::filter(centile != 1) %>%
  summarize(sprintf("%.100f", (1 - max(centile))))

qc_cent_df %>%
  summarize(sprintf("%.100f", min(centile)))
```
## CENTILES

@@

```{r}
#initalize empty dfs to store outputs
sum.df <- data.frame(pheno = character(),
                     term = character(),
                     df = double(),
                     sumsq = double(),
                     meansq = double(),
                     statistic = double(),
                     p.value = double())
t.df <- data.frame("pheno" = character(),
                   "group1" = character(),
                   "group2" = character(),
                   "p.value" = double())
pheno_diff.list <- paste0("abs.diff_", pheno_list)

#run tests
attach(cent_abs.diffs)
for (pheno in pheno_diff.list) {
  # df.aov <- tidy(kruskal.test(cent_abs.diffs[[pheno]] ~ cent_abs.diffs[["dataset"]]))
  # df.aov$pheno <- pheno
  # sum.df <- rbind(sum.df, df.aov)

  df.pairwise.t <- tidy(pairwise.rank.welch.t.test(x=cent_abs.diffs[[pheno]], g=cent_abs.diffs[["dataset"]], paired = TRUE, p.adj = "none"))
  df.pairwise.t$pheno <- pheno
  t.df <- rbind(t.df, df.pairwise.t)
}
detach()

@@

# t.df.sum <- t.df %>%
#   group_by(comp) %>%
#   summarize(n_sig_regions = sum(sig_fdr),
#             n_compare = n(),
#             min_pval = min(p.val_fdr),
#             max_pval = max(p.val_fdr))
# t.df.sum
```

Little plot
```{r}
#get means w/in feature 
mean.diffs.long <- cent_abs.diffs %>%
  mutate(dataset = factor(dataset, levels = c("ukb_CN", "cf", "cf.lm", "cf.gam", "cf.gamlss"), ordered = TRUE)) %>%
  group_by(dataset) %>%
  summarize(across(starts_with("abs.diff_"), mean)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="centile_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)),
      label = sub("_[^_]*_", "_", pheno))
```

@@

@@

```{r, eval=FALSE}
#calc errors
z_abs.diffs <- combined_preds %>%
  dplyr::filter(!str_detect(dataset, "refA")) %>% #drop ref site results
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref = "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(ends_with(".z"), ~ abs(. - first(.)), .names = "abs.diff_{.col}")) %>%
  ungroup() %>%
  dplyr::filter(dataset != "ukb_CN") #drop raw
saveRDS(z_abs.diffs, "data/z_abs.diffs.rds")
```

```{r}
z_abs.diffs <- readRDS("data/z_abs.diffs.rds")
```

@@

z.pheno_abs.diff.list <- paste0("abs.diff_", pheno_list, ".z")

#run tests
attach(z_abs.diffs)
for (pheno in z.pheno_abs.diff.list) {
  # df.aov <- tidy(kruskal.test(z_abs.diffs[[pheno]] ~ z_abs.diffs[["dataset"]]))
  # df.aov$pheno <- pheno
  # z.sum.df <- rbind(z.sum.df, df.aov)

  df.pairwise.t <- tidy(pairwise.rank.welch.t.test(x=z_abs.diffs[[pheno]], g=z_abs.diffs[["dataset"]], paired = TRUE, p.adj = "none"))
  df.pairwise.t$pheno <- pheno
  z.t.df <- rbind(z.t.df, df.pairwise.t)
}
detach()

@@

# sum_z.t.df <- z.t.df %>%
#   group_by(comp) %>%
#   summarize(n_sig_regions = sum(sig_fdr),
#             n_compare = n(),
#             min.pval = min(p.val_fdr),
#             max.pval = max(p.val_fdr))
# 
# sum_z.t.df
```

Little plot
```{r}
#get means w/in feature 
z.mean.diffs.long <- z_abs.diffs %>%
  mutate(dataset = factor(dataset, levels = c("ukb_CN", "cf", "cf.lm", "cf.gam", "cf.gamlss"), ordered = TRUE)) %>%
  group_by(dataset) %>%
  summarize(across(starts_with("abs.diff_"), mean)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="z_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno = gsub(".z", "", pheno)) %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)),
      label = sub("_[^_]*_", "_", pheno))
```

@@

## SCALE EFFECTS OF SEX
fix age, get diff between male and female pred. variance
```{r}
pred.csvs <- list.files(path = data_path, pattern = "_variance.csv", full.names = TRUE)
combined_var <- data.frame()

@@

##CENTILE FANS

@@

@@

Visualize
```{r}
sex_bias_ratio_abs.tests_df %>%
  dplyr::filter(sig_fdr == TRUE) %>%
  mutate(dataset = factor(dataset, levels = c("cf", "cf.lm", "cf.gam", "cf.gamlss"), ordered = TRUE),
         prop = factor(prop, levels=n_prop_list, ordered=TRUE)) %>%
  ggplot(aes(x=dataset, y=prop, fill=sex_diff)) + 
    geom_raster() +
    paletteer::scale_fill_paletteer_c("ggthemes::Classic Green-Blue") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    coord_fixed()
```
-log(p)

>>>>

# Setup 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(data.table, ggplot2, tidyverse, ggseg, ggsegExtra, paletteer, ggpubr, gridExtra, lemon, parallel, rstatix, lme4, interactions, ggrepel, ggrain)
#Sys.setenv("MC_CORES"=10L)
options("mc.cores")

@@

```{r}
data_path <- "/Users/megardn/Desktop/BGD_Repos/combat_biovar/derivatives"

@@

pheno_list.z <- paste0(pheno_list, ".z")
pheno_abs.diff.list <- paste0("abs.diff_", pheno_list)
z.pheno_abs.diff.list <- paste0("abs.diff_", pheno_list, ".z")

#also load raw data
ukb_cn_df <- fread("/Users/megardn/Desktop/BGD_Repos/combat_biovar/data/ukb_to_model/ukb_CN_data_simsites.csv")
```

@@

Within each combat config, test if there are significant differences in M and F subject's mean abs. centile error. Can test mean (not abs) centile errors, Z errors, etc, using `to_test` arg. FDR corrects for number of datasets tested

@@

  #apply fdr correction
  n_comp=length(unique(df$dataset))
  df.sex.t <- df.sex.t %>%
    dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = (n_comp*comp_multiplier)), #use comp_multiplier if using fun with lapply
           sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                               p.val_fdr >= 0.05 ~ FALSE))
  return(df.sex.t)
  #detach()
}

@@

  #conduct t test
  df.lm <- df %>%
    group_by(dataset) %>%
    #returns values for last (highest order) predictor
    summarise(p.value = tail(tidy(lm(as.formula(f)))$p.value, n=1),
              estimate = tail(tidy(lm(as.formula(f)))$estimate, n=1),
              t.stat = tail(tidy(lm(as.formula(f)))$statistic[2], n=1),
              std.error = tail(tidy(lm(as.formula(f)))$std.error[2]), n=1) %>%
    ungroup()
  n_comp=length(unique(df$dataset))

@@

pulled variance from cubic using `scp -r 'gardnerm@cubic-login.uphs.upenn.edu:/cbica/home/gardnerm/combat-biovar/ukb_basic/centile_csvs/*_variance.csv' .`

Get predicted centiles
- pulled from cubic using `scp -r 'gardnerm@cubic-login.uphs.upenn.edu:/cbica/home/gardnerm/combat-biovar/ukb_basic/centile_csvs/*_predictions.csv' .`

@@

```{r}
combined_preds %>% 
  filter_at(vars(pheno_list), any_vars(. %in% c(1, 0))) %>%
  dplyr::select(participant) %>%
  unique()

combined_preds %>% 
  filter_at(vars(pheno_list.z), any_vars(. %in% c(NA, Inf, -Inf))) %>%
  dplyr::select(participant) %>%
  unique()

# test.df <- combined_preds %>% 
#   dplyr::select(pheno_list)
# 
# test.df[test.df==1] <- 0
# 
# format(round(max(do.call(pmax, test.df)), 20), nsmall = 20)
```

saving out csv forces some centile scores to round up to 1, BUT it appears z scores are still being calculated correctly so just going to proceed!

Get errors
- pulled centile/z error calcs from cubic using `scp -r 'gardnerm@cubic-login.uphs.upenn.edu:/cbica/home/gardnerm/combat-biovar/ukb_basic/centile_csvs/*_diffs.csv' .`

```{r}
raw_files <- list.files(path = paste0(data_path, "/ukb_basic"), pattern = "_diffs.csv", full.names = TRUE) %>% unlist()

diffs.df <- fread(file = raw_files, sep=",") %>%
     mutate(dataset = gsub("_data|_data_predictions.csv|prop-|[0-9]|-", "", Source_File))
```

Get subject-wise summaries
- pulled from cubic using `scp -r 'gardnerm@cubic-login.uphs.upenn.edu:/cbica/home/gardnerm/combat-biovar/ukb_basic/centile_csvs/subject-wise/*.csv' .`

```{r}
raw_files <- list.files(path = paste0(data_path, "/ukb_basic"), pattern = "_subj_pred.csv", full.names = TRUE) %>% unlist()

subj_sum.df <- fread(file = raw_files, sep=",") %>%
     mutate(dataset = gsub("_data|_data_predictions.csv|prop-|[0-9]|-", "", Source_File))
```

## CENTILES

@@

```{r}
#initalize empty dfs to store outputs
sum.df <- data.frame(pheno = character(),
                     term = character(),
                     df = double(),
                     sumsq = double(),
                     meansq = double(),
                     statistic = double(),
                     p.value = double())
t.df <- data.frame("pheno" = character(),
                   "group1" = character(),
                   "group2" = character(),
                   "p.value" = double())

#run tests
attach(diffs.df)
for (pheno in pheno_diff.list) {
  # df.aov <- tidy(kruskal.test(diffs.df[[pheno]] ~ diffs.df[["dataset"]]))
  # df.aov$pheno <- pheno
  # sum.df <- rbind(sum.df, df.aov)

  df.pairwise.t <- tidy(pairwise.rank.welch.t.test(x=diffs.df[[pheno]], g=diffs.df[["dataset"]], paired = TRUE, p.adj = "none"))
  df.pairwise.t$pheno <- pheno
  t.df <- rbind(t.df, df.pairwise.t)
}
detach()

@@

t.df.sum <- t.df %>%
  group_by(comp) %>%
  summarize(n_sig_regions = sum(sig_fdr),
            n_compare = n(),
            min_pval = min(p.val_fdr),
            max_pval = max(p.val_fdr))
t.df.sum
```

Little plot
```{r}
#get means w/in feature 
mean.diffs.long <- diffs.df %>%
  dplyr::select(sex, dataset, pheno, pheno_abs.diff.list) %>%
  mutate(dataset = factor(dataset, levels = c("ukb_CN", "cf", "cf.lm", "cf.gam", "cf.gamlss"), ordered = TRUE)) %>%
  group_by(dataset) %>%
  summarize(across(starts_with("abs.diff_"), mean)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="centile_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)),
      label = sub("_[^_]*_", "_", pheno))
```

@@

### Sex-Biases

Visualize magnitude mean w/in subj error

```{r}
subj_sum.df %>%
  mutate(mean_cent_abs.diff = mean_cent_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_violin(aes(x=dataset, y=mean_cent_abs.diff, fill=sex)) +
  labs(y= "Magnitude Centile Error", x = "ComBat Configurations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean magnitude of error in centile scores across ComBat configurations")

subj_sum.df %>%
  mutate(mean_cent_diff = mean_cent_diff*100) %>% #for visuals
  ggplot(aes(x=dataset, y=mean_cent_diff, fill=sex)) +
  geom_boxplot(trim=FALSE, scale="width") +
  labs(y="Centile Error", x = "ComBat Configurations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean error in centile scores across ComBat configurations")

subj_sum.df %>%
  dplyr::filter(dataset == "cf.gamlss" | dataset == "cf.gam" | dataset == "cf.lm") %>%
  mutate(mean_cent_abs.diff = mean_cent_abs.diff*100) %>% #for visuals
  ggplot(aes(x=sim.site, y=mean_cent_abs.diff, fill=sex)) +
  geom_violin(trim=FALSE) +
  facet_wrap(~dataset) +
  labs(y="Magnitude Centile Error", x = "Simulated Site") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean magnitude error in centile scores across ComBat configurations")

subj_sum.df %>%
  ggplot() +
  geom_bar(aes(x=sim.site, fill=sex))
```

Significance testing (w/in sub. mean abs centile error) *ask about this - might be better to use wilcox since variances are ~= across sexes*

```{r}
sex.bias.t.tests(subj_sum.df)
```

@@

@@

#run tests
attach(diffs.df)
for (pheno in z.pheno_abs.diff.list) {
  # df.aov <- tidy(kruskal.test(diffs.df[[pheno]] ~ diffs.df[["dataset"]]))
  # df.aov$pheno <- pheno
  # z.sum.df <- rbind(z.sum.df, df.aov)

  df.pairwise.t <- tidy(pairwise.rank.welch.t.test(x=diffs.df[[pheno]], g=diffs.df[["dataset"]], paired = TRUE, p.adj = "none"))
  df.pairwise.t$pheno <- pheno
  z.t.df <- rbind(z.t.df, df.pairwise.t)
}
detach()

@@

sum_z.t.df <- z.t.df %>%
  group_by(comp) %>%
  summarize(n_sig_regions = sum(sig_fdr),
            n_compare = n(),
            min.pval = min(p.val_fdr),
            max.pval = max(p.val_fdr))

sum_z.t.df
```

Little plot
```{r}
#get means w/in feature 
z.mean.diffs.long <- diffs.df %>%
  dplyr::select(sex, dataset, pheno, z.pheno_abs.diff.list) %>%
  mutate(dataset = factor(dataset, levels = c("ukb_CN", "cf", "cf.lm", "cf.gam", "cf.gamlss"), ordered = TRUE)) %>%
  group_by(dataset) %>%
  summarize(across(starts_with("abs.diff_"), mean)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="z_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno = gsub(".z", "", pheno)) %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)),
      label = sub("_[^_]*_", "_", pheno))
```

@@

### Sex-Biases

Visualize magnitude mean w/in subj error

```{r}
subj_sum.df %>%
  ggplot() +
  geom_violin(aes(x=dataset, y=mean_z_abs.diff, fill=sex)) +
  labs(y= "Magnitude Z-Score Error", x = "ComBat Configurations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean magnitude error in z-scores across ComBat configurations")

subj_sum.df %>%
  ggplot(aes(x=sex, y=mean_z_diff, fill=sex)) +
  geom_boxplot(trim=FALSE, scale="width") +
  facet_wrap(~dataset, scale="free_y") +
  labs(y="Z-Score Error", x = "ComBat Configurations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean error in z-scores across ComBat configurations")

subj_sum.df %>%
  dplyr::filter(dataset == "cf.gamlss" | dataset == "cf.gam" | dataset == "cf.lm") %>%
  ggplot(aes(x=sim.site, y=mean_z_abs.diff, fill=sex)) +
  geom_violin(trim=FALSE) +
  facet_wrap(~dataset) +
  labs(y="Magnitude Z Error", x = "Simulated Site") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Within-subject mean magnitude error in z-scores across ComBat configurations")

subj_sum.df %>%
  ggplot() +
  geom_bar(aes(x=sim.site, fill=sex))
```

Significance testing (w/in sub. mean abs z error) *ask about this - might be better to use wilcox since variances are ~= across sexes*

```{r}
sex.bias.t.tests(subj_sum.df, to_test="mean_z_abs.diff")
sex.bias.t.tests(subj_sum.df, to_test="mean_z_diff")
```

## SCALE EFFECTS OF SEX

fix age, get diff between male and female pred. variance

```{r}
pred.csvs <- list.files(path = data_path, pattern = "_variance.csv", full.names = TRUE)
combined_var <- data.frame()

@@

## CENTILE FANS

@@

maybe better to get sex_summary.csv - just need significance

@@

Visualize
```{r}
sex_bias_ratio_abs.tests_df %>%
  dplyr::filter(sig_fdr == TRUE) %>%
  mutate(dataset = factor(dataset, levels = c("cf", "cf.lm", "cf.gam", "cf.gamlss"), ordered = TRUE),
         prop = factor(prop, levels=n_prop_list, ordered=TRUE)) %>%
  ggplot(aes(x=dataset, y=prop, fill=sex_diff)) + 
    geom_raster() +
    paletteer::scale_fill_paletteer_c("ggthemes::Classic Green-Blue") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    coord_fixed()
```
